\bf A note on the instability and degeneracy of deep learning models

### 2017-11-02 writing/paper-note/docs/note.tex
nordman's changes

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
\usepackage{graphicx,psfrag,epsf}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL

\usepackage{longtable,booktabs}
\usepackage[unicode=true]{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{tikz, subfig, amsthm, multirow, float, enumerate}
\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning}
\usepackage{bm}

%%INCLUDED-AK
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}

@@

\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }

%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

@@

We consider the behavior, and the potential impropriety, sequences of
discrete probability models built to incorporate observations of
increasing sample size \(N\). Interest is in identifying instability in
such models, which is roughly characterized by probabilities with
extreme sensitivity to small changes in data configuration. The concept
of instability was introduced in the field of statistical physics (i.e.,
point processes) by Ruelle
(\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further
extended by Schweinberger Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) for a
family of exponential models. At issue, models exhibiting instability
are typically undesirable as these tend to provide poor representations
of data or data-generation. As an example, such models can include
near-degenerate distributions that assign all essential probability mass
to only a subset of an overall sample space. The latter issue in
connection to degeneracy has been recognized as concern in that dominant
model outcomes may not resemble observed data (cf. Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). As a compounding
issue, model instability often has direct negative impacts for
statistical inference and computations based on likelihood functions.
Namely, volatilities in probability structure can potentially hamper the
numerical evaluations required for maximum likelihood estimation as well
as other model-based simulations via Markov Chain Monte Carlo (MCMC).
These reasons motivate our general study of instability for a broad
class of probability models, described next.

In the model framework, let \(\boldsymbol X_N = (X_1, \dots, X_N)\)
denote a collection of discrete random variables with a finite sample
space, \(\mathcal{X}^N\), represented as some \(N\)-fold Cartesian
product. That is, \(\mathcal{X}\) with \(|\mathcal{X}| < \infty\)
denotes the set of potential outcomes for each single variable \(X_i\),
so that the product space \(\mathcal{X}^N\) corresponds to values for
the variables \(\boldsymbol X_N=(X_1,\ldots,X_N)\). For each \(N\), let
\(P_{\boldsymbol \theta_N}\) denote a probability model on
\(\mathcal{X}^N\), under which
\(P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0\) is the probability of
the data outcome \((x_1, \dots, x_N) \in \mathcal{X}^N\). In this, we
assume that the model support of \(P_{\boldsymbol \theta_N}\) is the
sample space \(\mathcal{X}^N\). This framework produces probability
models \(P_{\boldsymbol \theta_N}\), indexed by a generic sequence of
parameters \(\boldsymbol \theta_N\), to describe data
\(\boldsymbol X_N\) of any given sample size \(N \geq 1\). For
simplicity, we will refer to this distributional class as \emph{Finite
Outcome Everywhere Supported (FOES)} models in the following. The
dimension and structure of such parameters are generic, without
restriction, though natural cases will be seen to include those where
\(\boldsymbol \theta_N \in \mathbb{R}^{q(N)}\) for some arbitrary
integer-valued function \(q(\cdot) \geq 1\).

Section \ref{examples} provides some examples of FOES models encountered
in graph/network analysis and machine learning (i.e., deep learning
models). These are used as references for later illustrations. Section
\ref{instability-results} then establishes several formal results for
FOES models with regard to instability. Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally
developed instability results framed for, or specific to, a certain
class of discrete exponential models. For similar exponential models
with random networks, Handcock
(\protect\hyperlink{ref-handcock2003assessing}{2003}) studied model
degeneracy, where a probability model places near complete mass on modes
and may thereby narrow the feasible model outcomes. As findings here and
from Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest,
model instability and degeneracy may also be related by viewing
degeneracy as an extreme, or limiting form, of instability. Our main
results establish a broad characterization of model instability,
appropriate across the whole FOES model class, that incorporates results
of Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a
special case. We prescribe a general and simple condition for
identifying instability in a FOES model sequence, which quantifies
whether certain maximal probabilities in a FOES model are too extreme
relative to the sample size \(N\). When fulfilled, the probability
structure of a FOES model is shown to exhibit extreme sensitivity, with
probability assignments possessing extreme peaks and troughs across
nearly identical outcomes. As the measure of model instability
increases, probabilities from an unstable FOES model additionally
increase in volatility and provably slide into degeneracy. Section
\ref{implications} then emphasizes the implications of such model
instability, showing that such impropriety can be expected to
numerically hinder maximum likelihood estimation and MCMC-based
simulations. As one potential remedy, suggestions are given or
constraining model parameterizations and avoid the most problematic
regions of the parameter space. Proofs of the main results appear in
Appendix \ref{appendix-instab}.

@@

\subsection{Discrete exponential family
models}\label{discrete-exponential-family-models}

For random variables
\(\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)\) with sample
space \(\mathcal{X}^N\), \(|\mathcal{X}| < \infty\), consider an
exponential family model for \(\boldsymbol X\) with probability mass
function given by
\begin{equation}
\label{eq:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector
\(\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}\) and natural
parameter function
\(\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L\) with fixed
positive integers \(k\) and \(L\) denoting their dimensions. Above,
\(\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L\) is a vector of
sufficient statistics, while \[
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
\] denotes the normalizing function with parameter space \(\Theta_N\).
The natural parameter function \(\eta (\boldsymbol \theta)\) has a
linear form (i.e.,
\(\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta\) for a given
\(L \times k\) matrix \(\bm{A}\)) in many common model formulaitons,
though may also be nonlinear (e.g., curved exponential families). In the
linear case, \(\eta (\boldsymbol \theta) = \boldsymbol \theta\) may be
generally assumed in the exponential parameterization with a minor
modification to the definition of sufficient statistics
\(\boldsymbol g_N(\boldsymbol x)\).

Such discrete exponential family models are special cases of the FOES
models, as seen by defining
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0\),
\(\boldsymbol x \in \mathcal{X}^N\), based on \eqref{eq:expo} and a
parameter sequence
\(\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k\). For example,
if observations \(\boldsymbol X = (X_1,\ldots,X_N)\) correspond to \(N\)
independent and identically distributed Bernoulli random variables, each
indicating a binary \(0\)-\(1\) outcome, the resulting probabilities
have exponential form \eqref{eq:expo} given by

@@

\label{eq:mod1}
P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, 
with sufficient statistic
\(\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N x_i\) and ``log odds
ratio'' parameter
\(\boldsymbol \theta_N \equiv \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}\).
More generally, supposing \(\boldsymbol X =(X_1,\ldots,X_N)\) represent
\(N\) independent trials, each assuming an outcome \(\{1,\ldots,k\}\)
among \(k\) possibilities (e.g., a die roll), a multinomial distribution
is given by

@@

\label{eq:mod11}
P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right] =
\exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, 
with sufficient statistic \(\boldsymbol g_N(\boldsymbol x)\) involving a
count \(\sum_{i=1}^N \mathbb{I}(x_i=j)\) for each outcome
\(j \in \{1,\ldots,k\}\), where \(\mathbb{I}(\cdot)\) denotes the
indicator function, and parameters
\(\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k\)
defining log-probability ratios
\(\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]\).
In addition to such standard models for discrete independent data,
exponential models of FOES type commonly arise with dependent spatial
data (Besag \protect\hyperlink{ref-besag1974spatial}{1974}) and
network/relational data (Wasserman and Faust
\protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For a random graph
or network with, say, \(n\) nodes, consider \(N={n \choose 2}\) random
edges where the \(i\)th edge is associated with a pair of nodes
\(s_i \equiv \{v_i,u_i\}\) and a binary variable \(X_i\in\{0,1\}\)
indicating presence/absence of an edge among the node pair \(s_i\),
\(i=1,\ldots,N\). Here the length \(N\) of the edge variable sequence
\(\boldsymbol X = (X_1,\ldots,X_N)\) increases as a function of node
number \(n\) and corresponding exponential models often incorporate
graph topographical features derived from \(\boldsymbol X\). As an
example, consider a graph model of exponential/FOES form prescribed by

@@

\label{eq:mod2}
P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,

@@

\] involving the numbers of edges, 2-stars and triangles among an
outcome \(\boldsymbol x\) given by \(g_{1,N}(\boldsymbol x)\),
\(g_{2,N}(\boldsymbol x)\) and \(g_{3,N}(\boldsymbol x)\), respectively,
along with \(k=3\) real parameters
\(\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})\).
For this network model \eqref{eq:mod2} in particular, as well as for more
general models of form \eqref{eq:expo}, Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered
instability in such exponential models with sequences of fixed
parameters
\(\boldsymbol \theta_N = (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k\),
\(N \geq 1\), of fixed dimension \(k\).

For model sequences
\(P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)\)
of the exponential type \eqref{eq:expo}, such as those in
\eqref{eq:mod1}-\eqref{eq:mod2}, note that the dimension \(k\) of the
parameter \(\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k\)
necessarily remains the same for all sample sizes \(N \geq 1\) as the
form of the natural parameter function \(\eta(\cdot)\) in \eqref{eq:expo}
and the number of sufficient statistics
\(\boldsymbol g_{N}(\boldsymbol x)\) do not depend on \(N\).
Consequently, \(\boldsymbol \theta_N\) lies in a parameter space of
fixed Euclidean dimension \(k\). However, this aspect need not be true
for other types of FOES models considered in Sections
\ref{restricted-boltzmann-machines} - \ref{deep-learning}, where instead
the numbers of parameters and sufficient statistics commonly increase
with the sample size \(N\).

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model
specified for discrete or continuous random variables, with binary
variables being most common (cf. Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}). A RBM
architecture has two layers, hidden (\(\mathcal{H}\)) and visible
(\(\mathcal{V}\)), with conditional independence within each layer. Let
\(\boldsymbol X = (X_1,\ldots,X_N)\) denote the \(N\) random variables
for visibles with support \(\mathcal{X}^N\) and
\(\boldsymbol H = (H_1,\ldots,H_{N_\mathcal{H}})\) denote the
\(N_\mathcal{H}\) random variables for hiddens with support
\(\mathcal{X}^{N_\mathcal{H}}\) where \(\mathcal{X} = \{-1,1\}\). For
parameters
\(\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}\),
\(\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N\), and
\(\boldsymbol \theta_N^{\mathcal{HV}}\) as a real matrix with dimension
\(N_\mathcal{H} \times N\), the RBM model for
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) has the joint
probability mass function

@@

\label{eq:RBM1}

@@

with normalizing function \[

@@

\] Let
\(\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}\),
with \(q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}\), denote the
parameter vector for the RBM, as indexed by the number \(N\) of visible
random variables (which may differ from the actual lengths of these
parameter vectors). The probability mass function for the visible
variables \(\boldsymbol X = (X_1, \dots, X_N)\) follows from
marginalizing the joint specification to yield
\label{eq:RBM2}

@@

Here the baseline model \eqref{eq:RBM1} for hidden/visible variables is a
linear exponential one in sufficient statistics
\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) using
\(\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)\) from
\eqref{eq:RBM1}, but the form differs from the previous exponential models
in \eqref{eq:expo} in that the lengths of parameters
\(\boldsymbol \theta_N\) and statistics
\((\tilde{\boldsymbol X}, \boldsymbol X^T\boldsymbol H)\) increase to
incorporate more visible variables. That is, in contrast to
\eqref{eq:expo}, the natural parameter function involved in the RBM model
\eqref{eq:RBM1}, as the identity mapping of the parameters
\(\boldsymbol \theta_N\in\mathbb{R}^{q(N)}\), naturally grows in
dimension \(q(N)\to \infty\) to accommodate visible variables
\(X_1, \dots, X_N\) of increasing sample size \(N\to \infty\).
Additionally, one may further arbitrarily choose the number
\(N_\mathcal{H}\) of hidden variables \(\boldsymbol H\) in the joint RMB
model \eqref{eq:RBM1} to define a marginal model \eqref{eq:RBM2} for the
\(N\) visible variables \(\boldsymbol X\), and the number
\(N_\mathcal{H}\) of hiddens may also potentially increase with \(N\).
Because \(|\mathcal{X}| = 2\) and
\(P_{\boldsymbol \theta_N}(\boldsymbol x) > 0\) for all
\(\boldsymbol x \in \mathcal{X}^N\), the RBM specification \eqref{eq:RBM2}
for visibles \(\boldsymbol\) corresponds to a FOES model, while the
joint distribution \eqref{eq:RBM1} for \((\boldsymbol X, \boldsymbol H)\)
is also a FOES model. As this example also indicates, any model formed
by marginalizing a base FOES model class, such as the RBM joint
specification \eqref{eq:RBM1}, is again a FOES model.

@@

@@

    based on $f_{\boldsymbol \theta_N}$ from \\eqref{eq:f} with components $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.

>>>>

\documentclass[12pt]{article}
\usepackage{amsmath,amssymb}
%\usepackage{graphicx,psfrag,epsf}
\usepackage{graphicx}
\usepackage{enumerate}
\usepackage{natbib}
\usepackage{url} % not crucial - just used below for the URL
\setlength{\textheight}{9in}
\usepackage{longtable,booktabs}
\usepackage[unicode=true]{hyperref}
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\usepackage{tikz,  amsthm, multirow, float}
%\usepackage{tikz-3dplot}
\usetikzlibrary{arrows, shapes, positioning}
\usepackage{bm}

\usepackage{color}
\newcommand{\red}[1]{{\color{red} #1}}
\newcommand{\blue}[1]{{\color{blue} #1}}
%%INCLUDED-AK
\usepackage{amsthm}
\newtheorem{theorem}{Theorem}
\newtheorem{proposition}{Proposition}
\newtheorem{lemma}{Lemma}

@@

\newcommand{\Gamc}{C_{N}(\boldsymbol \theta_N) }
\newcommand{\Gamt}{\Gamma_{N,2}(\boldsymbol \theta_N) }
%\pdfminorversion=4
% NOTE: To produce blinded version, replace "0" with "1" below.
\newcommand{\blind}{0}

@@

We consider the behavior, and the potential impropriety, sequences of discrete probability models built to incorporate observations of increasing sample size $N$.  Interest is in identifying  instability in such models, which is roughly characterized by probabilities with extreme sensitivity to small changes in data configuration.  The concept of instability was introduced in the field of statistical physics (i.e., point processes) by Ruelle (\protect\hyperlink{ref-ruelle1999statistical}{1999}) and then further extended by Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011})   for a family of exponential  models.
At issue, models exhibiting instability are typically undesirable  as these tend to provide poor representations of data or data-generation.
 As an example, such models  can include near-degenerate distributions that assign all essential probability mass to only a subset of
 an overall sample space.     The latter issue in connection to degeneracy has been recognized as concern  in that dominant model outcomes  may not resemble   observed data  (cf.~Handcock, \hyperlink{ref-handcock2003assessing}{2003}).  As a compounding issue, model instability often has direct negative impacts for statistical inference and computations based on likelihood functions.     Namely, volatilities  in probability structure  can potentially
  hamper the numerical evaluations required for maximum likelihood estimation as well as other model-based simulations via Markov Chain Monte Carlo (MCMC).
  These reasons motivate our general study of instability for a broad class of probability models, described next.

    In the model framework,
 let $  \boldsymbol X_N =(X_1, \dots, X_N)$ denote a collection of discrete random variables with a finite sample space, $\mathcal{X}^N$, represented as some $N$-fold Cartesian  product. That is, $\mathcal{X}$ with $|\mathcal{X}| < \infty$ denotes the set of potential outcomes for each single variable $X_i$, so that the product space $\mathcal{X}^N$ corresponds to values for the variables  $\boldsymbol X_N=(X_1,\ldots,X_N)$. For each $N$, let $P_{\boldsymbol \theta_N}$ denote a probability model on $\mathcal{X}^N$, under which $P_{\boldsymbol \theta_N}(x_1, \dots, x_N) > 0$ is the probability of the data outcome $(x_1, \dots, x_N) \in \mathcal{X}^N$. In this, we assume that the model support of $P_{\boldsymbol \theta_N}$ is the sample space $\mathcal{X}^N$. This framework produces probability models $P_{\boldsymbol \theta_N}$, indexed by a generic sequence of parameters $\boldsymbol \theta_N$, to describe data $\boldsymbol X_N$ of any given sample size $N \geq 1$. For simplicity, we will refer to this distributional class as {\it Finite Outcome Everywhere Supported} (FOES) models in the following.  The dimension and structure of such parameters are generic, without restriction, though natural cases will be seen to include those where $\boldsymbol \theta_N \in \mathbb{R}^{q(N)}$ for some arbitrary integer-valued function $q(\cdot) \geq 1$.

Section \ref{examples}   provides some examples of FOES models encountered in graph/network analysis and machine learning (i.e., deep learning models). These are used as references for later illustrations.  Section \ref{instability-results} then  establishes several formal results for  FOES models with regard to instability.  Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) originally developed instability results framed for, or specific to, a certain class of discrete exponential  models.  For similar exponential models with random networks,
Handcock (\protect\hyperlink{ref-handcock2003assessing}{2003})   studied model degeneracy, where a probability model places near complete mass on  modes and may thereby narrow the feasible
  model outcomes.    As findings here and from  Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) suggest, model instability and degeneracy may also be related by viewing degeneracy as an extreme, or limiting form, of instability.
Our main results establish a broad characterization of model instability, appropriate across the whole FOES model class,
 that incorporates results of
 Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) as a special case.
 We prescribe a general and simple condition  for identifying instability in a FOES model sequence,   which  quantifies whether certain maximal probabilities in a FOES model  are too extreme relative to the sample size $N$.
When fulfilled, the probability structure of a FOES model is shown to exhibit extreme sensitivity, with probability assignments possessing extreme  peaks and troughs across nearly identical outcomes.  As the measure of model instability increases,
probabilities from an unstable FOES model  additionally increase in volatility and provably slide into  degeneracy.
Section \ref{implications} then emphasizes the implications of such model instability, showing that such impropriety
can be expected to numerically hinder maximum likelihood estimation and   MCMC-based simulations.  As one potential remedy, suggestions are given
 for constraining model parameterizations and avoid the most problematic regions of the parameter space. Proofs of the main results appear in Appendix \ref{appendix-instab}.

@@

\subsection{Discrete exponential family models}
\label{discrete-exponential-family-models}

For  random variables $\boldsymbol X \equiv\boldsymbol X_N= (X_1, \dots, X_N)$ with sample space $\mathcal{X}^N$, $|\mathcal{X}| < \infty$, consider an exponential family model for $\boldsymbol X$ with probability mass function given by
\begin{equation}
\label{eqn:expo}
p_{N, \boldsymbol \theta}(\boldsymbol x) = \exp\left[\boldsymbol\eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) - \psi(\boldsymbol \theta)\right], \qquad \boldsymbol x \in \mathcal{X}^N,
\end{equation}
depending on parameter vector $\boldsymbol \theta \in \Theta_N \subset \mathbb{R}^{k}$ and natural parameter function $\boldsymbol \eta : \mathbb{R}^k \mapsto \mathbb{R}^L$ with fixed positive integers $k$ and $L$ denoting their dimensions. Above, $\boldsymbol g_N : \mathcal{X}^N \mapsto \mathbb{R}^L$ is a vector of sufficient statistics, while
$$
\psi(\boldsymbol \theta) = \log \sum\limits_{\boldsymbol x \in \mathcal{X}^N}\exp\left[\boldsymbol \eta^T(\boldsymbol \theta) \boldsymbol g_N(\boldsymbol x) \right], \qquad \boldsymbol \theta \in \Theta_N\equiv \{\boldsymbol \theta \in \mathbb{R}^k : \psi(\boldsymbol \theta) < \infty \},
$$
denotes the normalizing function with   parameter space $\Theta_N$.   The natural parameter function $\eta (\boldsymbol \theta) $  has a linear form (i.e., $\eta (\boldsymbol \theta)= \bm{A} \boldsymbol \theta$ for a given $L \times k$ matrix $\bm{A}$) in many common model formulaitons, though may also be nonlinear (e.g., curved exponential families).  In the linear case, $\eta (\boldsymbol \theta) =\boldsymbol \theta$ may be generally assumed in the exponential parameterization with a minor modification to the definition of sufficient statistics $\boldsymbol g_N(\boldsymbol x)$.

Such discrete exponential family models are special cases of the FOES models, as seen by  defining $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)> 0$,  $\boldsymbol x \in \mathcal{X}^N$, based on (\ref{eqn:expo}) and a parameter sequence $\boldsymbol \theta_N \in \Theta_N \subset \mathbb{R}^k$.  For example, if observations $\boldsymbol X=(X_1,\ldots,X_N)$ correspond to $N$ independent and identically distributed Bernoulli random variables, each indicating a binary $0$-$1$ outcome,  the resulting probabilities have exponential form (\ref{eqn:expo})  given by

@@

\label{eqn:mod1} P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\boldsymbol \theta_N \sum_{i=1}^N x_i\right], \qquad \boldsymbol x=(x_1,\ldots,x_N) \in\{0,1\}^N, \end{equation} with sufficient statistic $\boldsymbol g_N(\boldsymbol x)\equiv \sum_{i=1}^N  x_i$ and ``log odds ratio" parameter $\boldsymbol \theta_N \equiv  \log[ P_{\boldsymbol \theta_N}(X_i=1)/P_{\boldsymbol \theta_N}(X_i=0) ] \in \mathbb{R}$.  More generally,
 supposing $\boldsymbol X =(X_1,\ldots,X_N)$ represent  $N$
independent trials, each assuming an outcome $\{1,\ldots,k\}$ among $k$ possibilities (e.g., a die roll), a multinomial distribution is given by

@@

   \label{eqn:mod11} P_{\boldsymbol \theta_N}(\boldsymbol x) \propto  \exp\left[  \boldsymbol \theta_{N}^T g_N(\boldsymbol x)   \right]=
  \exp\left[ \sum_{j=1}^k {\theta_{j,N}} \sum_{i=1}^N \mathbb{I}(x_i=j) \right], \qquad \boldsymbol x  \in\{1,\ldots,k\}^N, \end{equation}
with sufficient statistic $\boldsymbol g_N(\boldsymbol x)$ involving a count $\sum_{i=1}^N \mathbb{I}(x_i=j)$  for each outcome $j\in\{1,\ldots,k\}$, where $\mathbb{I}(\cdot)$ denotes the indicator function,   and parameters $\boldsymbol \theta_N=(\theta_{1,N},\ldots,\theta_{k,N})\in\mathbb{R}^k$ defining log-probability ratios
$\theta_{i,N}-\theta_{j,N} =\log [P_{\boldsymbol \theta_N}(X_1=i)/P_{\boldsymbol \theta_N}(X_1=j)]$.  In addition to such standard models for discrete independent data,
exponential models of FOES type commonly arise with
 dependent spatial data   (Besag
\protect\hyperlink{ref-besag1974spatial}{1974}) and network/relational data (Wasserman
and Faust \protect\hyperlink{ref-wasserman1994social}{1994}; Handcock
\protect\hyperlink{ref-handcock2003assessing}{2003}). For a random graph  or network  with, say, $n$ nodes,  consider $N={n \choose 2}$ random edges
 where the $i$th edge is associated with a pair of nodes $s_i \equiv \{v_i,u_i\}$  and a binary variable $X_i\in\{0,1\}$  indicating   presence/absence of an edge among the   node pair $s_i$, $i=1,\ldots,N$.  Here the length $N$ of the edge variable sequence $\boldsymbol X =(X_1,\ldots,X_N)$ increases as a function of node number $n$ and corresponding
 exponential models often incorporate graph topographical features derived from $\boldsymbol X$. As an example,  consider a   graph model of exponential/FOES form prescribed by

@@

\label{eqn:mod2} P_{\boldsymbol \theta_N}(\boldsymbol x) \propto
 \exp\left[\sum_{j=1}^3 \theta_{j,N} g_{j,N}(\boldsymbol x)\right], \quad\qquad \boldsymbol x=(x_1,\ldots,x_N)  \in\{0,1\}^N,\end{equation}

@@

  \]
 involving the numbers of edges, 2-stars and triangles among an outcome $\boldsymbol x$ given by   $ g_{1,N}(\boldsymbol x)$,   $ g_{2,N}(\boldsymbol x)$  and  $ g_{3,N}(\boldsymbol x)$, respectively, along   with $k=3$ real parameters $\boldsymbol \theta_N \equiv (\theta_{1,N},\theta_{2,N},\theta_{3,N})$.   For this network model (\ref{eqn:mod2}) in particular, as well as for more general models of form (\ref{eqn:expo}),  Schweinberger
(\protect\hyperlink{ref-schweinberger2011instability}{2011}) considered instability in such exponential models   with sequences of fixed parameters $\boldsymbol \theta_N= (\theta_1,\ldots,\theta_k)\in\mathbb{R}^k$, $N \geq 1$, of fixed dimension $k$.

  For model sequences $P_{\boldsymbol \theta_N}(\boldsymbol x)\equiv p_{N,\boldsymbol \theta_N}(\boldsymbol x)$
  of the exponential type (\ref{eqn:expo}), such as those in (\ref{eqn:mod1})-(\ref{eqn:mod2}),
   note that the dimension $k$ of the parameter $\boldsymbol \theta_N\in\Theta \subset \mathbb{R}^k$  necessarily remains the same for all sample sizes $N \geq 1$ as the form of the natural parameter function $\eta(\cdot)$ in (\ref{eqn:expo}) and the number of sufficient statistics $\boldsymbol g_{N}(\boldsymbol x)$ do not depend on $N$.  Consequently,
   $\boldsymbol \theta_N$ lies in a parameter space of fixed Euclidean dimension $k$.  However, this aspect need not be true for other types of FOES models considered in Sections \ref{restricted-boltzmann-machines}-\ref{deep-learning}, where instead the numbers of parameters and sufficient statistics commonly increase with the sample size $N$.

%\subsection{Discrete exponential family
%models}\label{discrete-exponential-family-models}

@@

A restricted Boltzmann machine (RBM) is an undirected graphical model specified for discrete or continuous random variables, with binary variables being most common (cf.~Smolensky
\protect\hyperlink{ref-smolensky1986information}{1986}).  A RBM architecture has two layers, hidden ($\mathcal{H}$) and visible ($\mathcal{V}$), with conditional independence within each layer. Let $\boldsymbol X=(X_1,\ldots,X_N)$ denote the $N$ random variables for visibles with support $\mathcal{X}^N$ and $\boldsymbol H=(H_1,\ldots,H_{N_\mathcal{H}})$ denote the $N_\mathcal{H}$ random variables for hiddens with support $\mathcal{X}^{N_\mathcal{H}}$ where  $\mathcal{X} = \{-1,1\}$. For parameters $\boldsymbol \theta_N^{\mathcal{H}} \in \mathbb{R}^{N_\mathcal{H}}$, $\boldsymbol \theta_N^{\mathcal{V}}\in \mathbb{R}^N$, and $\boldsymbol \theta_N^{\mathcal{HV}}$ as a real matrix with dimension $N_\mathcal{H} \times N$, the RBM model for $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$  has the joint probability mass function

@@

\label{eqn:RBM1}

@@

with normalizing function

@@

 Let $\boldsymbol \theta_N = (\boldsymbol \theta_N^{\mathcal{H}}, \boldsymbol \theta_N^{\mathcal{V}}, \boldsymbol\theta_N^{\mathcal{HV}} ) \in \Theta_N \equiv \mathbb{R}^{q(N)}$, with $q(N) = N + N_\mathcal{H} + N*N_\mathcal{H}$, denote the parameter vector  for the RBM,  as indexed by the number $N$
of visible random variables (which may differ from the actual lengths of these parameter vectors).    The probability mass function for the visible variables $\boldsymbol X =(X_1, \dots, X_N)$ follows from marginalizing the joint specification to yield
\begin{equation}\label{eqn:RBM2}

@@

Here the baseline model (\ref{eqn:RBM1}) for hidden/visible variables is a linear exponential one in sufficient statistics  $(\tilde{\boldsymbol X},  \boldsymbol X^T\boldsymbol H)$ using $\tilde{\boldsymbol X}=(\boldsymbol X,\boldsymbol H)$ from   (\ref{eqn:RBM1}), but the form differs from the previous exponential models
in (\ref{eqn:expo}) in that the lengths of parameters $\boldsymbol \theta_N$ and   statistics  $(\tilde{\boldsymbol X},  \boldsymbol X^T\boldsymbol H)$ increase to incorporate more visible variables.  That is, in contrast to (\ref{eqn:expo}), the natural parameter function involved in the RBM model (\ref{eqn:RBM1}), as the identity mapping of the parameters $\boldsymbol \theta_N\in\mathbb{R}^{q(N)}$,    naturally grows in  dimension  $q(N)\to \infty$   to accommodate visible variables $X_1, \dots, X_N$ of increasing sample size $N\to \infty$.  Additionally, one may further arbitrarily choose the number $N_\mathcal{H}$ of hidden variables $\boldsymbol H$ in the joint RMB model (\ref{eqn:RBM1}) to define a marginal model (\ref{eqn:RBM2}) for the $N$ visible variables $\boldsymbol X$, and the number $N_\mathcal{H}$ of hiddens may also potentially   increase with $N$. Because   $|\mathcal{X}| = 2$ and $P_{\boldsymbol \theta_N}(\boldsymbol x) > 0$ for all $\boldsymbol x \in \mathcal{X}^N$, the RBM  specification (\ref{eqn:RBM2}) for visibles $\boldsymbol$ corresponds to a FOES model, while the joint distribution (\ref{eqn:RBM1}) for $(\boldsymbol X, \boldsymbol H)$ is also a FOES model.  As this example also indicates,  any model formed by marginalizing a base FOES model class, such as the RBM joint specification (\ref{eqn:RBM1}), is again a FOES model.

@@

\]

@@

\]
based on $f_{\boldsymbol \theta_N}$ from (\ref{eqn:f}) with components
 $\boldsymbol x \in \{\pm 1\}^{N}, \boldsymbol h \in \{\pm 1\}^{N_{\mathcal{H}}}$.  \\
